Resonant Scattering in a Uniform Sphere with Large Optical Depth

### 2021-09-21 ms.tex
Incorporating Phil's edits up to the time-dependent diffusion section

\subsection{Comparison to Monte Carlo}
The Monte Carlo method is used to solve the transfer equation numerically. This method is valid at arbitrarily large optical depths, being restricted only by its computational demand which grows in proportion to the number of photons used. For each simulation, a total of $2^{20} \approx 10^6$ photon packets are initialized at a monochromatic source frequency $x_s$ and are allowed to propagate through the medium until escaping, at which point their positions, outgoing angles, and escape frequencies are stored. A constant temperature of $T=10000\ \rm K$ is set for the gas. Frequency redistribution is calculated at each scattering, including the effects of recoil, to obtain the spectrum at the surface of the spherical simulation domain. In the comparisons shown in this section, the raw photon data is binned in frequency to obtain spectra. 64 bins are used to ensure good statistics and minimal noise in each bin. Further details of the Monte Carlo implementation are discussed in \cite{2017ApJ...851..150H}.

@@

We now compare each of the previously-discussed solutions for surface flux with Monte Carlo simulations. The escape probability $P(x)$ is defined as the total luminosity at the surface divided by the source luminosity, or
\be
P(x) = \frac{16\pi^2R^2H(R, x)\Delta}{L}
\ee
This is defined such that $\int P(x)dx = 1$. Since $H(r, \nu)$ is implicitly written per $d\nu$, a factor of $\Delta$ gives the expression the correct units. In Fig. \ref{fig:sol_mc_residual_0}, the Monte Carlo escape probability as a function of frequency is shown with that of the solutions $H_{\rm d}$, $H_{\rm 0}$, and $H_{\rm 0 + bc} = H_{\rm 0} + H_{\rm bc}$ for an optical depth of $\tau_0 = 10^7$ and photons emitted at line center $\rm x_s = 0$.  The solution with the correct frequency-dependent boundary condition enforced, $H_{\rm 0 + bc}$, has considerably lower residuals to Monte Carlo than the other solutions, especially in the line wing. It is evident that the divergent solution $H_{\rm d}$ fails in the line wings. The $H_{\rm bc}$ term is negative at the peak of the spectrum and positive in the line wing such that, when added to the solution $H_0$ derived by \citet{2006ApJ...649...14D}, it adequately corrects for the apparent excess of flux in the peaks of the spectrum. This peak excess is also present in previous work that used the \citet{1990ApJ...350..216N} solution, such as \citet{2015MNRAS.449.4336S} where the discrepancy is particularly noticeable for line center optical depths of $\tau_0=10^5$ and $10^6$. The boundary term resolves the slight deficit of $H_{\rm0}$ in the line wings, further improving agreement with the numerical result. Note that the residuals to the $H_0$ solution are a close match to the $H_{\rm bc}$ term, since the Monte Carlo represents the ``true'' solution, $H$, and $H_{\rm bc} = H - H_0$.

 \begin{figure}
    \centering
    \includegraphics{tau_threepanel.pdf}
    \caption{The same spectrum as in Fig. \ref{fig:sol_mc_residual_0}, but for three different line center optical depths $\tau_0 = 10^5, 10^6, 10^7$ with the $x$ and $y$ axes scaled by the characteristic frequency diffusion scale $(a\tau_0)^{1/3}$. Photons are initialized with $\rm x = 0.0$ in each case.} 
    \label{fig:sol_mc_tau}
\end{figure}

The size of the frequency-dependent boundary condition correction is dependent on the line center optical depth, which also dictates the x-position of the peak of spectrum according to Eq. \ref{eq:tau_peak_scaling}. This term contributes a significant amount of flux correction even at line center optical depths of $10^7$ where the $J=0$ solution is expected to perform its best, i.e., the majority of photons are pushed further out into the wing and thus the simplifying assumptions made in the solution of the differential equation are better founded. One factor of $(a\tau_0)^{1/3}$ has been scaled out of the x-axis of Fig. \ref{fig:sol_mc_tau} such that the peaks of the distributions are vertically aligned. This scaling has also been applied to the y-axis to preserve normalization of the escape probability. From Eq. \ref{eq:hbc_scaling}, the size of the term $H_{\rm bc}$ should become smaller with larger optical depths, following a $(a\tau_0)^{-1/3}$ scaling. Indeed, agreement between both analytic flux calculations and the Monte Carlo points in Fig. \ref{fig:sol_mc_tau} improves with $H_{\rm bc}$ providing a fractionally smaller correction to $H_0$ as the optical depth becomes larger. At lower optical depths, the scattering of photons within the Doppler core of the line becomes important, yet our solution is not constructed to model this behavior precisely. The effects of core scattering can be seen in the Monte Carlo data, however, as the peaks of the distribution shift substantially outward in frequency from the line core for $\tau_0 = 10^5$.

@@

Since the solutions have been generalized to allow for a monochromatic source of photons at a frequency $\sigma_s$ away from line center, their agreement with Monte Carlo as a function of source frequency can also be characterized. Photons initialized closer to the line wing may travel larger mean free paths as the opacity is sensitively dependent on the line profile. The larger spatial diffusion contributes to greater escape probability, biasing the frequency distribution of escaping photons. In the limit that $|\rm x_s|$ becomes large, the distribution becomes a delta function in frequency as all photons escape the gas without scattering. Fig. \ref{fig:sol_mc_xinit} shows calculations performed for three initial photon frequencies $\rm x_s = 0.0, 6.0$, and $12.0$ for a line center optical depth of $\tau_0=10^7$. The bias in frequency distribution is slight for $\rm x_s = 6.0$ where the line profile is still relatively large and emitted photons remain trapped in the core, but the increase in mean free path is much more substantial when photons are initialized with $\rm x_s=12.0$, further away from the line core. It is seen here that the discrepancy between the Monte Carlo data and $H_0$ (effectively the scale of the corrective $H_{\rm bc}$ term) becomes larger as $\rm x_s$ increases. Thus, for sufficiently large $\tau_0$ and emission away from line center $x_s \neq 0$, the importance of enforcing frequency-dependent boundary conditions at the surface of the sphere grows. The bias in the escaping spectrum is substantial at lower optical depths, shown in Fig. \ref{fig:sol_mc_xinit_lowtau}, where a $\Delta x_s$ of 6 is a larger fraction of the spectral peak location $(a\tau_0)^{1/3}$. It is clear from the figure that as $x_s$ becomes further from the optically thick core, the spectrum approaches a delta function in frequency at $\sigma_s$. Additionally, since the material is optically thin here, we expect there to be a stronger disagreement between the Monte Carlo and the analytic solution as derivations assumed large optical depths.

@@

The previous figures show that as $\tau_0$ increases, $H_0$ appears to become more accurate. There is agreement between the Monte Carlo and $H_0$ when $H_{\rm bc}$ is small which motivates a quantitative examination of how the boundary term $H_{\rm bc}$ scales with both $\tau_0$ and $\sigma_s$, shown in Fig. \ref{fig:xinit}. In this graphic, the photon initialization frequency $\sigma_s$ is shifted in integer multiples of $\tau_0$ such that $\sigma_s$ falls near the peak of the spectrum for each $\tau_0$. For clarity, only the $\sigma > 0$ side of the spectrum is shown. From Eq. \ref{eq:hbc_scaling}, it is expected that the size of the correction $H_{\rm bc}$ should become smaller with larger optical depths following a $(a\tau_0)^{-1/3}$ scaling. This factor has been scaled out of the figure such that solutions for different optical depths and the same source frequency $\sigma_s$ should show close agreement in scale on the figure's vertical axis if the relation holds. Indeed, the scaled solutions converge as the optical depth becomes larger, indicating general agreement with the $(a\tau_0)^{1/3}$ scaling. The remaining discrepancy present in the vertical axis for fixed $\sigma_s$ is due to the width of the Doppler core with respect to the peak of the spectrum; at lower depths, this causes the assumptions about the functional form of the line profile to break down. From this, we conclude that the errors introduced by the incorrect separation of variables in \cite{1973MNRAS.162...43H}, \cite{1990ApJ...350..216N}, \cite{2006ApJ...649...14D} and others are proportional to $(a\tau_0)^{1/3}$, where $a$ is the ratio of natural to Doppler broadening and $\tau_0$ is the optical depth at line center. Furthermore, an additional amount of error of order $(a\tau_0)^{1/3}$ is introduced as the photon frequency $\sigma_s$ from line center is increased by a factor of $\tau_0$.

>>>>

\subsection{Comparison to Monte Carlo}
The Monte Carlo method is used to solve the transfer equation numerically. This method is valid at arbitrarily large optical depths, being restricted only by its computational demand which grows proportionally to the number of photons used and $\tau_0$. For each simulation, a total of ${\sim}10^6$ photon packets are initialized at a monochromatic source frequency $x_s$ and are allowed to propagate through the sphere until escaping, at which point their positions, outgoing angles, and escape frequencies are stored. A constant temperature of $T=10,000\ \rm K$ is set for the gas. Frequency redistribution is calculated at each scattering, including the effects of recoil, to obtain the spectrum at the surface of the spherical simulation domain. In the comparisons shown in this section, the raw photon data is binned in frequency to obtain spectra. 64 bins are used to ensure good statistics and minimal noise in each bin. Further details of the Monte Carlo implementation are discussed in \cite{2017ApJ...851..150H}.

@@

We now compare each of the previously-discussed solutions for surface flux with Monte Carlo simulations. We return to the Doppler units $x$ for the photon frequency rather than the $\sigma$ units used in the solutions of the transfer equation. The spectrum $P(x)$ is defined as the specific luminosity at the surface divided by the source luminosity, or
\be
P(x) = \frac{16\pi^2R^2H(R, x)\Delta}{L}.
\ee
This is normalized such that $\int P(x)dx = 1$. Since $H(r, x)$ is per $d\nu$, a factor of $\Delta$ gives the expression the correct units.

% TODO: Edits unclear. Organization of sentences OK? "Describe what's in the figure and then discuss" - Mention of Hd moved to end of paragraph, details of the Hbc term moved higher up.
In Figure \ref{fig:sol_mc_residual_0}, the Monte Carlo escape probability as a function of frequency is shown with that of the solutions $H_{\rm d}$, $H_{\rm 0}$, and $H_{\rm 0 + bc} = H_{\rm 0} + H_{\rm bc}$ for an optical depth of $\tau_0 = 10^7$ and photons emitted at line center $\rm x_s = 0$.  The $H_{\rm bc}$ term is negative at the peak of the spectrum and positive in the line wing such that, when added to the solution $H_0$ derived by \citet{2006ApJ...649...14D}, it adequately corrects for the apparent excess of flux in the peaks of the spectrum. This peak excess is also present in previous work that used the \citet{1990ApJ...350..216N} solution, such as \citet{2015MNRAS.449.4336S} where the discrepancy is particularly noticeable for line center optical depths of $\tau_0=10^5$ and $10^6$ in the top panel of their Figure 5. The solution with the correct frequency-dependent boundary condition enforced, $H_{\rm 0 + bc}$, has considerably lower residuals to Monte Carlo than the other solutions, especially in the line wing. The boundary term corrects the slight deficit of $H_{\rm0}$ in the line wings, further improving agreement with the numerical result. Note that the residuals to the $H_0$ solution are a close match to the $H_{\rm bc}$ term, since the Monte Carlo represents the ``true'' solution, $H$, and $H_{\rm bc} = H - H_0$. It is evident that the divergent solution $H_{\rm d}$ fails in the line wings.

 \begin{figure}
    \centering
    \includegraphics{tau_threepanel.pdf}
    \caption{The same as Figure \ref{fig:sol_mc_residual_0}, but for $\tau_0 = 10^5, 10^6, 10^7$. The $x$ and $y$ axes are scaled by $(a\tau_0)^{1/3}$. Photons are initialized with $\rm x_s = 0.0$ in each case.} 
    \label{fig:sol_mc_tau}
\end{figure}

The size of $H_{\rm bc}$ is dependent on $\tau_0$, which also dictates the x-position of the peak of spectrum according to Eq. \ref{eq:tau_peak_scaling}. This term is significant even at $\tau_0 {\sim} 10^7$ where $H_0$ solution is expected to perform its best, i.e., photons are pushed further out into the wing and thus the simplifying assumptions made in the solution of the differential equation are a better approximation.

In Figure \ref{fig:sol_mc_tau} we show the solutions alongside Monte Carlo for $\tau_0=10^5, 10^6$, and $10^7$. From Eq. \ref{eq:hbc_scaling}, the size of the term $H_{\rm bc}$ should become smaller with larger optical depths, following a $(a\tau_0)^{-1/3}$ scaling. Indeed, agreement between Eq. \ref{eq:H0surf} and the Monte Carlo points in Fig. \ref{fig:sol_mc_tau} improves as $\tau_0$ increases, with $H_{\rm bc}$ providing a fractionally smaller correction to $H_0$. One factor of $(a\tau_0)^{1/3}$ has been scaled out of the x-axis such that the peaks of the distributions are vertically aligned. This scaling has also been applied to the y-axis to preserve normalization of the escape probability. At lower $\tau_0$, the scattering of photons within the Doppler core of the line becomes important, yet our solution does not include this effect. The effects of line core scattering can be seen in the Monte Carlo data, however, as the peaks of the distribution shift substantially outward in frequency from the line core for $\tau_0 = 10^5$.

@@

Since the solutions have been generalized to allow for a monochromatic source of photons at a frequency $x_s \neq 0$, their agreement with Monte Carlo as a function of source frequency can also be characterized. Photons initialized further out in the line wing have larger mean free paths. The larger spatial diffusion contributes to greater escape probability for these photons. In the limit that $|\rm x_s|$ becomes large, the distribution becomes a delta function in frequency as all photons escape the gas without scattering.

Figure \ref{fig:sol_mc_xinit} shows calculations performed for three initial photon frequencies $\rm x_s = 0.0, 6.0$, and $12.0$ for a line center optical depth of $\tau_0=10^7$. The asymmetry of the spectrum is slight for $\rm x_s = 6.0$ where the cross section is still relatively large, but the increase in mean free path is much more substantial when photons are initialized with $\rm x_s=12.0$, further away from the line core. It is seen here that the difference between the Monte Carlo data and $H_0$ becomes larger as $\rm x_s$ increases. Thus, for sufficiently large $\tau_0$ and emission away from line center ($x_s \neq 0$), the importance of enforcing frequency-dependent boundary conditions at the surface of the sphere grows.

Figure \ref{fig:sol_mc_xinit_lowtau} shows emission away from line center at the same values of $x_s$ as in Figure \ref{fig:sol_mc_xinit}, but for $\tau_0=10^6$ rather than $10^7$. The difference between the left and right side of the escaping spectrum is substantial at lower optical depths, since a change in $x_s$ of 6 is a larger fraction of the spectral peak location $(a\tau_0)^{1/3}$ for lower $\tau_0$. It is clear from the figure that as $x_s$ becomes further from the line core, the spectrum approaches a delta function in frequency. Additionally, since the material is optically thin here, we expect there to be a stronger disagreement between the Monte Carlo and the analytic solution as derivations assumed large optical depths.

@@

Figure \ref{fig:xinit} shows quantitative examination of how the boundary term $H_{\rm bc}$ scales with both $\tau_0$ and $x_s$, motivated by the previous figures' demonstration that $H_0$ appears to become more accurate as $\tau_0$ increases. In this figure, the photon initialization frequency $x_s$ is shifted by integer multiples of $\tau_0$ in $\sigma$ frequency units (Eq. \ref{app:rteqn_derivation}\ref{eq:change_of_variables}) such that the source, now written $\sigma_s$, falls near the peak of the spectrum for each $\tau_0$. For clarity, only the $x > 0$ side of the spectrum is shown. From Eq. \ref{eq:hbc_scaling}, it is expected that the size of $H_{\rm bc}$ should become smaller with larger optical depths following a $(a\tau_0)^{-1/3}$ scaling as compared to $H_0$. This factor has been scaled out of the figure such that solutions for different optical depths and the same source frequency $\sigma_s$ should show close agreement in scale on the figure's vertical axis if the relation holds. Indeed, the scaled solutions converge as the optical depth becomes larger, indicating general agreement with the $(a\tau_0)^{-1/3}$ scaling. The remaining discrepancy present in the vertical axis for fixed $\sigma_s$ is due to the width of the Doppler core with respect to the peak of the spectrum; at lower $\tau_0$, this causes the line profile approximation in the wing (Eq. \ref{app:rteqn_derivation}\ref{eq:line_profile_wing}) to break down. From this, we conclude that the errors introduced by the incorrect separation of variables in \cite{1973MNRAS.162...43H}, \cite{1990ApJ...350..216N}, \cite{2006ApJ...649...14D} and others are proportional to $(a\tau_0)^{-1/3}$.
